Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-26 projectReport/chapter5.tex
Found too many small errors

In this chapter we turn to the implementation of the function $G_T$ by discussing a method to learn the function $\Gamma$. We apply our methodology to some chaotic attractors. 
Most notably, we consider the reconstruction of the chaotic attractor of the double pendulum -- an  attractor that has not previously been successfully reconstructed because of its intricate structure. 
In fact the previous data-driven approaches also have not been successful in building a model from the double pendulum.

@@

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\eqref{eqns_from_data}\ednote{How do I get both?}. To reach this point though, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value, lies within the range $[-1,1]$, and has been zero-padded as required (see paragraph below for further discussion), we drive the system with external input through the function $g$.

@@

We implement a feedforward network (a network with no memory, unlike the RNN) to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library .

@@

Once $\Gamma$ is learnt, we progress to the the prediction step where $u_n$ is forecasted from $(x_{n-1},x_{n})$. Note that  due to causal embedding $(x_{n-1},x_{n})$ is in one-to-one correspondence with the left-infinite orbit $(\ldots,u_{n-1},u_{n-1})$. 
When $u_n$ is computed using $\Gamma$, it is fed to the driven system $g$ along with  $x_n$ to obtain $x_{n+1}$. Now we apply $\Gamma$ on $(x_n,x_{n+1})$ to forecast $u_{n+1}$ and the whole process repeats. Note that this corresponds to implementing the two model equations for learning from data~\eqref{eqns_from_data}.

@@

We now turn our attention to another practical issue related to the forecasting of chaotic dynamical systems.  
Recall that the twin equations~\ref{eqns_from_data} generate a model from data. 
For a chaotic system, a small amount of noise from even a very accurate model would cause solutions with the same initial conditions to diverge. This noise could be due either to input-noise or computational error.
Statistically, however,  a chaotic system will behave well in the sense that the visiting frequency (the average time spent by an orbit in any region of the phase space of the dynamical system) is identical for almost all initial conditions. 
The phrase `almost all' here is specified as a measure and refers to a result in ergodic theory (\cite{wikiErgodicTheory}) which is a topic beyond the scope of this project.
To illustrate the accuracy of the twin  equations~\ref{eqns_from_data} in modelling the data, we plot the densities of the trajectories of the actual and predicted solutions in ~\ref{fig:dp_success_density},~\ref{fig:noisydp_success_density} and~\ref{fig:Thomas_TrajDensity}.

In our implementation, the input and reservoir  matrices $A$ and $B$ in the RNN are sparse matrices. This sparsity is not essential but seems to perform slightly empirically better than choosing the matrices arbitrarily.  We also discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opt to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of 1. The condition number of a matrix $M$ is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed.  We do not have any theoretical justifications for this.

Lastly, we ensure the matrix $B$ has unit spectral radius so that the variable $\alpha$ in the RNN equation \eqref{eqn_driving} can be used to know if the USP is satisfied. 
It is a  feature that a matrix with a spectral radius less than 1 generally has the USP~\cite{yildiz2012re}.

@@

More explicitly, denote $X_{1:N}$ as the matrix with the first $N$ states of the network data represented by row vectors. 
$X_{1:N}=U\Sigma P^T$ here represents the singular value decomposition of $X_{1:N}$ ($U$ and $V$ are unitary matrices, and $\Sigma$ is a diagonal matrix) and we refer to \cite{wikiSVdecomp}. 
We then denote the principal component matrix by the symbol $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P$.
These principal components are used to train a feedforward neural network.

@@

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See Figure~\ref{fig:dp_setup} below.)
This system shows chaos in a striking manner -- consider once more Figure~\ref{fig:dp_sdic}.
The system has four variables: for each pendulum rod an associated angle and angular velocity. An ideal system is assumed where the rods are massless and pivots are frictionless.
\begin{figure}[ht]
  \includegraphics[width=0.8\linewidth]{Graphs/_dp_setup.eps}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP where $\theta, m, l$ represent the angle, rod-length, and pendulum head mass respectively.}
\label{fig:dp_setup}
\end{figure}

@@

% The second-order system is converted to a first-order ODE system and solved using the Runge-Kutta numerical method.  
Here we chose the values $m_1=2.0$, $m_2=2.0$, $L_1=1.5$ and $m_2=1.5$ with initial velocities $v_1=1.5, v_2=-2$ for the respective pendulum heads.
With a change of variables the second-order system can be written as a system of first order equations with four variables, where each pendulum has an associated $x$ and $y$ coordinates of the mass (the lengthy equations are not shown here). 
The ODE system is then solved using the fourth-order Runge-Kutta numerical method.

% Undamped DP data-sets were generated and used in this project due to the requirement that the map $T$ be surjective that we must not forget. 
% An example of this would be the case of a DP exhibiting energy-loss (i.e. where the map $T$ is not surjective). In such a case, the attractor is a single point in the phase space when the double pendulum comes to rest. 
% It must be mentioned once again that if we use data outside the attractor, learning is not reliable. 
% To see this, consider Figure~\ref{fig:damped_pendulum} and~\ref{fig:dp_notsurjective}.

@@

Having spent ample time to discuss the challenges, we now change tone so as to satisfy the optimist and present a successful run of the DP-data to predict the future movement of the pendulum.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_success_4coords_traj.eps} 
  \captionof{figure}{Predicted (red) and actual (blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads. From top to bottom are as follows: $x$-coordinate of topmost pendulum head, $y$-coordinate of topmost pendulum head, $x$-coordinate of lower pendulum head, $y$-coordinate of lower pendulum head. 
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 300 timesteps and stay close up until about 600 seconds.  } 
  \label{fig:dp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_success_4coords_hist.eps}
  \captionof{figure}{Densities of predicted and actual values of $x$- and $y$-coordinate of pendulum heads.}
  \label{fig:dp_success_density}
 \end{figure}

@@

For the signal-to-noise ration we adopt the formula $\text{SNR}_{dB}=10\text{log}_{10}\big(\frac{P_\text{{signal}}}{P_{\text{noise}}}\big)$ where $P_\text{{signal}}$ refers to the variance observed in the normalised data and $P_{\text{noise}}$ represents the variance of the added noise.
Once again we display the actual and predicted trajectories for the noisy DP data in Figure~\ref{fig:noisydp_success_traj}. 
Next we consider the densities of the predicted and actual values in Figure~\ref{fig:noisydp_success_density} and observe that they display highly similar distributions.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.95\linewidth]{Graphs/_dp_noise_FigCD_2030.eps} 
  \captionof{figure}{Predicted (red) and actual (blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads with noise added from a normal distribution with mean zero and standard deviation 0.1, equating to approximately 39dB's of noise.
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 250 timesteps and stay close up until about 400 seconds.  } 
 \label{fig:noisydp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_noise_2B_2030.eps}
  \captionof{figure}{Densities of predicted and actual values of x-coordinate of first pendulum head.}
  \label{fig:noisydp_success_density}
 \end{figure}

@@

\subsubsection{Thomas' Cyclically Symmetric Strange Attractor}\label{Thomas_Attractor}
Thomas' Cyclically Symmetric Strange Attractor, a 3D attractor proposed in 1999 by Ren\'e Thomas in~\cite{ThomasAttractor}, is described by a set of three equations which is distinctly reminiscent of another deterministic three-equation model exhibiting chaotic behaviour proposed by Lorenz. 
It has a single parameter $\beta$ and has been shown to transition to chaotic behavious when $\beta<0.208186$~\cite{Thomas_BetaParameter}.
\begin{eqnarray}\label{eqns_thomas}
  x_{n+1} = \sin(y_n) - \beta{x_n} \\
  y_{n+1} = \sin(z_n) - \beta{y_n} \\
  z_{n+1} = \sin(x_n) - \beta{z_n}.
\end{eqnarray}

@@

A sequence of 3000 observations in both the clear and noisy data-sets respectively was fed into the system with 300 entries discarded to simulate the network's loss of memory. 
The next 9000 steps were predicted and compared with true data. See Figure~\ref{fig:Thomas_trajectories} for the actual and predicted trajectories of the noise-free and perturbed systems respectively.

\begin{figure}[ht]
  \centering
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_noisy_0.1056.eps}
  \endminipage\hfill
  \minipage{0.5\linewidth}
    \centering    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_clean_0.1056.eps}
  \endminipage
 \captionof{figure}{True (red) and predicted (blue) trajectories for the Thomas attractor with parameter value $\beta=0.1056$.Left: Noise added from normal distribution with mean zero and standard deviation 0.05. Right: No noise added to dataset.}
\label{fig:Thomas_trajectories}
\end{figure}

@@

A finaldiscrete-time map named the Fractal Dream Attractor -- more commonly as the Pickover or Clifford Map (first discovered by Clifford A. Pickover and discussed in his fascinating book ``Chaos in Wonderland" \cite{PickoverChaos}) -- was also considered.
\begin{eqnarray}\label{eqns_clifford}
  {x_{n+1}=\sin(ay_n) + c\cos(ax_n)} \\
  {y_{n+1}=\sin(bx_n)+d\cos(by_n)}.
\end{eqnarray}

The behaviour of the system with a parameter values of $x_0=0$, $y_0=0$, $a = -1.7$,  $b = 1.8$, $c = -1.9$ and $d = -0.4$ was examined. 
Data was scaled to fit inside the interval $[-1,1]$. Only the noise-free case was investigated. 
Depictions of the empirical results are available in Figure~\ref{fig:Clifford}.
 A sequence of 25000 observations was fed into the system with 15000 entries discarded. 
 The reason for such a great number of discarded entries is the following: the system seems to have an incredibly long transient;driving it for less than 5000 entries only gives an accurate prediction to about 5 seconds and it starts improving when the discard is in the range $[10000,15000]$ to simulate the network's loss of memory. The next 1000 steps were predicted and compared with true data.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,left]{Graphs/_Clifford_1_nonoise.eps}
  \caption*{These graphs were constructed by predicting the Clifford system 1000 steps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. As perceived here, we are able to lock on to the trajectory of the Clifford map almost exactly for the first 25 steps.}
  \minipage{0.5\textwidth}
      \centering
      \includegraphics[width=\linewidth]{Graphs/_Clifford_2_nonoise.eps}
  \endminipage\hfill
  \minipage{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{Graphs/_Clifford_3_nonoise.eps}
  \endminipage
  \caption{The bottom row, left, shows a plot of the true attractor (red) and the predicted values (blue). Bottom row, right, shows the distribution of the first coordinate of the learnt system (blue) and the actual Clifford system (red). }
  \label{fig:Clifford}
\end{figure}

@@

Here we consider only the non-invertible map of the Clifford attractor to supplement the results in \cite{Supp}. 
This statistic is a generalisation of the Pearson correlation coefficient for real-valued random variables to random vectors (e.g., \cite{puccetti2019measuring}) and we present the details of computing this coefficient between $(x_{n-1},x_n)$ and $G_T(x_{n-1},x_n)$. 
If $\Sigma_{a}$ and $\Sigma_{b}$ denote the covariance matrices of the vectors $(x_{n-1},x_n)$ and $G_T(x_{n-1},x_n)$ respectively, and $\Sigma_{ab}$ denotes the covariance matrix between the vectors $(x_{n-1},x_n)$ and $G_T(x_{n-1},x_n)$, then the multidimensional correlation coefficient is computed by using the traces of these matrices.
\[
    \rho= \frac{\text{tr}({\Sigma_{ab}})}{\text{tr}({\sqrt{\Sigma_a\Sigma_b}})}.
\]
The value $\rho$ also satisfies one of the desired and well-known properties of the one-dimensional Pearson coefficient; that is, $\rho=\pm 1$ if and only if $Y\overset{d}{=}AX+b$ for some invertible matrix $A$ and vector $b$. 
In practice then, $\rho$, can be used to measure the linear relationship between two random vectors of the same dimension and thus serves as an indicator of functional complexity. 
For the non-invertible system defined by the Clifford attractor, we indicate the difference by which $\rho$ changes for $G_T$ in Table~\ref{tbl_attractorsPearson}.
Following~\cite{manjunath2021universal}, we state that the multidimensional correlation coefficient $\rho$ evidences a general reduction in the functional complexity of the map $G_T$ emerging due to a RNN. 
In Table~\ref{tbl_attractorsPearson}, the second column corresponds to the amount of artificial neurons (the dimension of $X$) in the implemented RNN.

@@

           \bottomrule
                \end{tabular}}\label{Table_FC}%\ednote{http://paulbourke.net/fractals/peterdejong/}
     \caption{
              Following~\cite{manjunath2021universal}, we determine the multidimensional correlation coefficients $\rho$ to evidence a general reduction in the functional complexity of the map $G_T$ emerging due to a RNN. Rows correspond to distinct dynamical systems considered in the experiments
              The second column corresponds to the amount of artificial neurons (the dimension of $X$) in the implemented RNN.
              The last two columns comprise numerical estimates of $\rho$ for the relevant vectors. The linear association between $u_n$ vs $T(u_{n})$, as well as the linear relationship 
              $\begin{bmatrix}
                x_{n-1}\\
                x_n
            \end{bmatrix}$ vs 
            $G_T 
            \left({\begin{bmatrix}
                x_{n-1}\\
                x_n
            \end{bmatrix}}\right)$ 
              may be contrasted by the reader. 
              Finally, the RNN's dimension (from \eqref{Seq_RNN})
              is provided to exhibiting that an increase in dimension of the driven system will 
              typically result in a map $G_T$ with less functional complexity.}\label{tbl_attractorsPearson}
      \end{table}
% \end{center}

>>>>

In this chapter we turn to the implementation of the function $G_T$ by discussing a method to learn the function $\Gamma$. We apply our methodology to some chaotic attractors. 
Most notably, we consider the reconstruction of the chaotic attractor of the double pendulum -- an  attractor that has not previously been successfully reconstructed because of its intricate structure. 
In fact, the previous data-driven approaches also have not been successful in building a model from the double pendulum.

@@

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\eqref{eqns_from_data}. To reach this point though, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value, lies within the range $[-1,1]$, and has been zero-padded as required (see paragraph below for further discussion), we drive the system with external input through the function $g$.

@@

We implement a feedforward network (a network with no memory, unlike the RNN) to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library.

@@

Once $\Gamma$ is learnt, we progress to the prediction step where $u_n$ is forecasted from $(x_{n-1},x_{n})$. Note that  due to causal embedding $(x_{n-1},x_{n})$ is in one-to-one correspondence with the left-infinite orbit $(\ldots,u_{n-1},u_{n-1})$. 
When $u_n$ is computed using $\Gamma$, it is fed to the driven system $g$ along with  $x_n$ to obtain $x_{n+1}$. Now we apply $\Gamma$ on $(x_n,x_{n+1})$ to forecast $u_{n+1}$ and the whole process repeats. Note that this corresponds to implementing the two model equations for learning from data~\eqref{eqns_from_data}.

@@

We now turn our attention to another practical issue related to the forecasting of chaotic dynamical systems.  
Recall that the twin equations~\ref{eqns_from_data} generate a model from data. 
For a chaotic system, a small amount of noise from even a very accurate model would cause solutions with the same initial conditions to diverge. This noise could be due either to input-noise or computational error.
Statistically, however,  a chaotic system will behave well in the sense that the visiting frequency (the average time spent by an orbit in any region of the phase space of the dynamical system) is identical for almost all initial conditions. 
The phrase `almost all' here is specified as a measure and refers to a result in ergodic theory (\cite{walters2000introduction}) which is a topic beyond the scope of this project.
To illustrate the accuracy of the twin  equations~\ref{eqns_from_data} in modelling the data, we plot the densities of the trajectories of the actual and predicted solutions in~\ref{fig:dp_success_density},~\ref{fig:noisydp_success_density} and~\ref{fig:Thomas_TrajDensity}.

In our implementation, the input and reservoir  matrices $A$ and $B$ in the RNN are sparse matrices. This sparsity is not essential but seems to perform slightly empirically better than choosing the matrices arbitrarily.  We also discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opt to enforce the requirement that both matrices be well-conditioned, i.e.,that they have a condition number of 1. The condition number of a matrix $M$ is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed.  We do not have any theoretical justifications for this.

Lastly, we ensure the matrix $B$ has unit spectral radius so that the variable $\alpha$ in the RNN equation~\eqref{eqn_driving} can be used to know if the USP is satisfied. 
It is a  feature that a matrix with a spectral radius less than 1 generally has the USP~\cite{yildiz2012re}.

@@

More explicitly, denote $X_{1:N}$ as the matrix with the first $N$ states of the network data represented by row vectors. 
$X_{1:N}=U\Sigma P^T$ here represents the singular value decomposition of $X_{1:N}$ ($U$ and $V$ are unitary matrices, and $\Sigma$ is a diagonal matrix) and we refer to~\cite{walters2000introduction}. 
We then denote the principal component matrix by the symbol $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P$.
These principal components are used to train a feedforward neural network.

@@

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See Figure~\ref{fig:dp_setup} below.)
This system shows chaos in a striking manner -- consider once more Figure~\ref{fig:dp_sdic}.
The system has four variables: for each pendulum rod an associated angle and angular velocity. An ideal system is assumed where the rods are massless, and pivots are frictionless.
\begin{figure}[ht]
  \includegraphics[width=0.4\linewidth]{Graphs/_dp_setup.eps}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP where $\theta, m, l$ represent the angle, rod-length, and pendulum head mass respectively.}
\label{fig:dp_setup}
\end{figure}

@@

% The second-order system is converted to a first-order ODE system and solved using the Runge-Kutta numerical method.  
Here we chose the values $m_1=2.0$, $m_2=2.0$, $L_1=1.5$ and $m_2=1.5$ with initial velocities $v_1=1.5, v_2=-2$ for the respective pendulum heads.
With a change of variables, the second-order system can be written as a system of first order equations with four variables, where each pendulum has an associated $x$ and $y$ coordinates of the mass (the lengthy equations are not shown here). 
The ODE system is then solved using the fourth-order Runge-Kutta numerical method.

% Undamped DP datasets were generated and used in this project due to the requirement that the map $T$ be surjective that we must not forget. 
% An example of this would be the case of a DP exhibiting energy-loss (i.e.,where the map $T$ is not surjective). In such a case, the attractor is a single point in the phase space when the double pendulum comes to rest. 
% It must be mentioned once again that if we use data outside the attractor, learning is not reliable. 
% To see this, consider Figure~\ref{fig:damped_pendulum} and~\ref{fig:dp_notsurjective}.

@@

Having spent ample time to discuss the challenges, we now change tone to satisfy the optimist and present a successful run of the DP-data to predict the future movement of the pendulum.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_success_4coords_traj.eps} 
  \captionof{figure}{Predicted (red) and actual (blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads. From top to bottom are as follows: $x$-coordinate of topmost pendulum head, $y$-coordinate of topmost pendulum head, $x$-coordinate of lower pendulum head, $y$-coordinate of lower pendulum head. 
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 300 timesteps and stay close until about 600 seconds.  } 
  \label{fig:dp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_success_4coords_hist.eps}
  \captionof{figure}{Densities of predicted and actual values of $x$- and $y$-coordinate of pendulum heads.}
  \label{fig:dp_success_density}
 \end{figure}

@@

For the signal-to-noise ration we adopt the formula $\text{SNR}_{dB}=10\text{log}_{10}\big(\frac{P_\text{{signal}}}{P_{\text{noise}}}\big)$ where $P_\text{{signal}}$ refers to the variance observed in the normalised data and $P_{\text{noise}}$ represents the variance of the added noise.
Once again, we display the actual and predicted trajectories for the noisy DP data in Figure~\ref{fig:noisydp_success_traj}. 
Next, we consider the densities of the predicted and actual values in Figure~\ref{fig:noisydp_success_density} and observe that they display highly similar distributions.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.95\linewidth]{Graphs/_dp_noise_FigCD_2030.eps} 
  \captionof{figure}{Predicted (red) and actual (blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads with noise added from a normal distribution with mean zero and standard deviation 0.1, equating to approximately 39dB's of noise.
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 250 timesteps and stay close until about 400 seconds.  } 
 \label{fig:noisydp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_noise_2B_2030.eps}
  \captionof{figure}{Densities of predicted and actual values of x-coordinate of first pendulum head.}
  \label{fig:noisydp_success_density}
 \end{figure}

@@

\subsubsection{Thomas' Cyclically Symmetric Strange Attractor}\label{Thomas_Attractor}
Thomas' Cyclically Symmetric Strange Attractor, a 3D attractor proposed in 1999 by Ren\'e Thomas in~\cite{ThomasAttractor}, is described by a set of three equations which is distinctly reminiscent of another deterministic three-equation model exhibiting chaotic behaviour proposed by Lorenz. 
It has a single parameter $\beta$ and has been shown to transition to chaotic behaviour when $\beta<0.208186$~\cite{Thomas_BetaParameter}.
\begin{eqnarray}\label{eqns_thomas}
  x_{n+1} = \sin(y_n) - \beta{x_n} \\
  y_{n+1} = \sin(z_n) - \beta{y_n} \\
  z_{n+1} = \sin(x_n) - \beta{z_n}.
\end{eqnarray}

@@

A sequence of 3000 observations in both the clear and noisy datasets respectively was fed into the system with 300 entries discarded to simulate the network's loss of memory. 
The next 9000 steps were predicted and compared with true data. See Figure~\ref{fig:Thomas_trajectories} for the actual and predicted trajectories of the noise-free and perturbed systems respectively.

\begin{figure}[ht]
  \centering
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_noisy_0.1056.eps}
  \endminipage\hfill
  \minipage{0.5\linewidth}
    \centering    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_clean_0.1056.eps}
  \endminipage
 \captionof{figure}{True (red) and predicted (blue) trajectories for the Thomas attractor with parameter value $\beta=0.1056$. Left: Noise added from normal distribution with mean zero and standard deviation 0.05. Right: No noise added to dataset.}
\label{fig:Thomas_trajectories}
\end{figure}

@@

A finaldiscrete-time map named the Fractal Dream Attractor -- more commonly as the Pickover or Clifford Map (first discovered by Clifford A. Pickover and discussed in his fascinating book ``Chaos in Wonderland"~\cite{PickoverChaos}) -- was also considered.
\begin{eqnarray}\label{eqns_clifford}
  {x_{n+1}=\sin(ay_n) + c\cos(ax_n)} \\
  {y_{n+1}=\sin(bx_n)+d\cos(by_n)}.
\end{eqnarray}

The behaviour of the system with parameter values of $x_0=0$, $y_0=0$, $a = -1.7$,  $b = 1.8$, $c = -1.9$ and $d = -0.4$ was examined. 
Data was scaled to fit inside the interval $[-1,1]$. Only the noise-free case was investigated. 
Depictions of the empirical results are available in Figure~\ref{fig:Clifford}.
 A sequence of 25000 observations was fed into the system with 15000 entries discarded. 
 The reason for such a great number of discarded entries is the following: the system seems to have an incredibly long transient; driving it for less than 5000 entries only gives an accurate prediction to about 5 seconds and it starts improving when the discard is in the range $[10000,15000]$ to simulate the network's loss of memory. The next 1000 steps were predicted and compared with true data.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,left]{Graphs/_Clifford_1_nonoise.eps}
  \caption*{These graphs were constructed by predicting the Clifford system 1000 steps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. As perceived here, we can lock on to the trajectory of the Clifford map almost exactly for the first 25 steps.}
  \minipage{0.5\textwidth}
      \centering
      \includegraphics[width=\linewidth]{Graphs/_Clifford_2_nonoise.eps}
  \endminipage\hfill
  \minipage{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{Graphs/_Clifford_3_nonoise.eps}
  \endminipage
  \caption{The bottom row, left, shows a plot of the true attractor (red) and the predicted values (blue). Bottom row, right, shows the distribution of the first coordinate of the learnt system (blue) and the actual Clifford system (red). }
  \label{fig:Clifford}
\end{figure}

@@

Here we consider only the non-invertible map of the Clifford attractor to supplement the results in~\cite{Supp}. 
This statistic is a generalisation of the Pearson correlation coefficient for real-valued random variables to random vectors (e.g.,~\cite{puccetti2019measuring}) and we present the details of computing this coefficient between $(x_{n-1},x_n)$ and $G_T(x_{n-1},x_n)$. 
If $\Sigma_{a}$ and $\Sigma_{b}$ denote the covariance matrices of the vectors $(x_{n-1},x_n)$ and $G_T(x_{n-1},x_n)$ respectively, and $\Sigma_{ab}$ denotes the covariance matrix between the vectors $(x_{n-1},x_n)$ and $G_T(x_{n-1},x_n)$, then the multidimensional correlation coefficient is computed by using the traces of these matrices.
\[
    \rho= \frac{\text{tr}({\Sigma_{ab}})}{\text{tr}({\sqrt{\Sigma_a\Sigma_b}})}.
\]
The value $\rho$ also satisfies one of the desired and well-known properties of the one-dimensional Pearson coefficient: that is, $\rho=\pm 1$ if and only if $Y\overset{d}{=}AX+b$ for some invertible matrix $A$ and vector $b$. 
In practice then, $\rho$, can be used to measure the linear relationship between two random vectors of the same dimension and thus serves as an indicator of functional complexity. 
For the non-invertible system defined by the Clifford attractor, we indicate the difference by which $\rho$ changes for $G_T$ in Table~\ref{tbl_attractorsPearson}.
Following~\cite{manjunath2021universal}, we state that the multidimensional correlation coefficient $\rho$ evidences a general reduction in the functional complexity of the map $G_T$ emerging due to a RNN. 
In Table~\ref{tbl_attractorsPearson}, the second column corresponds number of artificial neurons (the dimension of $X$) in the implemented RNN.

@@

           \bottomrule
                \end{tabular}}\label{Table_FC}%\ednote{http://paulbourke.net/fractals/peterdejong/}
     \caption{
              Following~\cite{manjunath2021universal}, we determine the multidimensional correlation coefficients $\rho$ to evidence a general reduction in the functional complexity of the map $G_T$ emerging due to a RNN. Rows correspond to distinct dynamical systems considered in the experiments
              The second column corresponds number of artificial neurons (the dimension of $X$) in the implemented RNN.
              The last two columns comprise numerical estimates of $\rho$ for the relevant vectors. The linear association between $u_n$ vs $T(u_{n})$, as well as the linear relationship 
              $\begin{bmatrix}
                x_{n-1}\\
                x_n
            \end{bmatrix}$ vs 
            $G_T 
            \left({\begin{bmatrix}
                x_{n-1}\\
                x_n
            \end{bmatrix}}\right)$ 
              may be contrasted by the reader. 
              Finally, the RNN's dimension (from~\eqref{Seq_RNN})
              is provided to exhibiting that an increase in dimension of the driven system will 
              typically result in a map $G_T$ with less functional complexity.}\label{tbl_attractorsPearson}
      \end{table}
% \end{center}

### 2022-01-26 projectReport/chapter6.tex
Found too many small errors

Each discrete-time state space model has at its core a driven dynamical system. 
In this project, it has been observed that an input originating from a compact space entailing exactly one solution (i.e., if the driven system has the USP) is equivalent to representing information without distortion, a concept which was formalised by establishing the existence of a causal mapping. 
The established results are general, implying that any driven dynamical system with specific properties, such as SI-invertibility, can give rise to a causal embedding. 
In particular, we have a topological conjugacy between the inverse-limit space of the underlying dynamical system and the single-delay lag dynamics of a driven system~\cite{manjunath2013echo}.
Finite-length input data is shown to be sufficient for forecasting as the left-infinite segment of the past of the input beyond some arbitrarily large but finite time (a result which holds due to the continuity of the universal semi-conjugacy) influences the driven system's states in an almost negligibly small manner.

@@

Finally, we have considered some physical systems to observe the methodology in practice and verify that we can indeed predict the future evolution of chaotic systems such as the double pendulum.
We demonstrated the robustness of our methodology to external noise by considering the cases of the  Dream Attractor and the Thomas' attractor.
\ednote{A sufficient ending?}
%\ednote{Loose paragraph commented out}
% Considered from a philosophical-scientific  perspective, this project has aimed to establish a juncture between rationalism (the theory of nonautonomous dynamical systems) and empiricism (machine learning) to yield results profitable to the discipline of Applied Mathematics. 
% This was done by showing that dynamics in an observed space can be established as topologically conjugate to the underlying system. 
% As a result of the topological conjugacy, deep learning techniques are then subsequently applied to physically such equations from the data. 
% In so doing, we acquire models of high fidelity that provide exceptional forecasting results on chaotic data that exhibit both statistical and topological consistency in the long run.

>>>>

Each discrete-time state space model has at its core a driven dynamical system. 
In this project, it has been observed that an input originating from a compact space entailing exactly one solution (i.e., if the driven system has the USP) is equivalent to representing information without distortion, a concept which was formalised by establishing the existence of a causal mapping. 
The established results are general, implying that any driven dynamical system with specific properties, such as SI-invertibility, can give rise to a causal embedding. 
We have a topological conjugacy between the inverse-limit space of the underlying dynamical system and the single-delay lag dynamics of a driven system~\cite{manjunath2013echo}.
Finite-length input data is shown to be sufficient for forecasting as the left-infinite segment of the past of the input beyond some arbitrarily large but finite time (a result which holds due to the continuity of the universal semi-conjugacy) influences the driven system's states in an almost negligibly small manner.

@@

Finally, we have considered some physical systems to observe the methodology in practice and verify that we can indeed predict the future evolution of chaotic systems such as the double pendulum.
We demonstrated the robustness of our methodology to external noise by considering the cases of the  Fractal Dream Attractor and the Thomas' attractor.

%\ednote{Loose paragraph commented out}
% Considered from a philosophical-scientific  perspective, this project has aimed to establish a juncture between rationalism (the theory of nonautonomous dynamical systems) and empiricism (machine learning) to yield results profitable to the discipline of Applied Mathematics. 
% This was done by showing that dynamics in an observed space can be established as topologically conjugate to the underlying system. 
% As a result of the topological conjugacy, deep learning techniques are then subsequently applied to physically such equations from the data. 
% In so doing, we acquire models of high fidelity that provide exceptional forecasting results on chaotic data that exhibit both statistical and topological consistency in the long run.
